---
title: "Bedrocan Cannabis Supplemental Data Analysis"
author: "Sophie Watts"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_notebook:
    code_folding: hide
    theme: cerulean
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}


#load packages
library("readr")
library("dplyr")
library("RColorBrewer")
library("gplots")
library("viridis")
library("scales")
library("tidyverse")
library("cowplot")
library("gridExtra")
library("grid")
library("ggplot2")
library("rstatix")
library("vegan")
library("patchwork")

theme_set(theme_classic())

```
#Boxplots of chemical concentration with label type
```{r}

chem_data <- read_csv("../outputs/chem_data.csv")

chem_data$type <- as.factor(chem_data$type)

#select phenos
chem_data <- chem_data %>% select(type:canna_total)

p <- list()
for (i in 2:ncol(chem_data)) {
  chemical_dat <- chem_data[, c(1,i)]
  name <- colnames(chemical_dat)[2]
  colnames(chemical_dat)[2] <- "chem"
  p[[i]] <-ggplot(chemical_dat, aes(x=type, y=chem, fill=type)) + 
  geom_boxplot(alpha=0.7)+
  ylab(paste(name, " (mg/g)"))+
  xlab("Sativa-Indica labelling scale")+
  scale_fill_manual(values = c("#feb24c", "#fd8d3c","#fc4e2a","#bd0026", "#800026"))+
  theme(legend.position="none")
}

p <-p[-1]
multi_page <- ggarrange(plotlist = p , nrow = 4, ncol = 2)
multi_page[[1]]
ggexport(multi_page, filename = "../figures/chem_label_boxplots.pdf")

```

#Mantel test.

```{r mantel, warning=FALSE, message=FALSE}
#compare the matrices of genetic distances vs. chemical distances using a mantel test.

#load 1-IBS matrix that was made with PLINK
ibs_gen <- read_delim("../data/20191209_bedrocan_gen_filtered.mdist", 
                                "\t", escape_double = FALSE, col_names = FALSE, 
                                trim_ws = TRUE)


#load sample names for mdist file
ibs_gen_samples <- read_delim("../data/20191209_bedrocan_gen_filtered.mdist.id", 
                                      "\t", escape_double = FALSE, col_names = FALSE, 
                                      trim_ws = TRUE)

#name the col and rows with sample names
rownames(ibs_gen) <- ibs_gen_samples$X1
colnames(ibs_gen) <- ibs_gen_samples$X2

#load chemical data
gen_chem_data <- read_csv("../outputs/gen_chem_data.csv")

#make vector of ibs_gen_samples
samples_vec <- ibs_gen_samples$X1

#rearrange gen_chem_data by sample order from samples_vec
gen_chem_data <- gen_chem_data %>% slice(match(samples_vec, Sample))
chem_mat_euclid <- gen_chem_data[,6:45]
#calculate euclidean distances between all samples and make matrix
eucl.chem<-as.matrix(dist(chem_mat_euclid, method="euclidean"))
#rename col and rownames with sample names.
rownames(eucl.chem)<-gen_chem_data$Sample
colnames(eucl.chem)<-gen_chem_data$Sample

ibs_gen <- as.matrix(ibs_gen)

#perform mantel test.
mantel(ibs_gen, eucl.chem)

gen_vec <- as.vector(ibs_gen[upper.tri(ibs_gen)])
chem_vec <- as.vector(eucl.chem[upper.tri(eucl.chem)])

#merge both distance vectors
distances <- data.frame(gen_vec, chem_vec, stringsAsFactors=FALSE)

#plot
mantel <- ggplot(distances, aes(gen_vec, chem_vec))+
  geom_point(size=1, alpha=0.7, colour = "darkgrey")+
  theme_bw()+
  labs(x="Genetic distance", y="Chemical distance")+
  theme(panel.border = element_blank(), axis.text=element_text(colour = "black", size=10), axis.title=element_text(size = 10, face = "bold"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black", size = 0.2), plot.title = element_text(size = 9))+
  ggtitle("mantel r = 0.21 p-value = 1x10-3")
#ggsave("../figures/mantel.pdf", plot = mantel)

mantel

```

#Chemical distance between clones.

```{r chem_clones, message=FALSE, warning=FALSE}

#Chemical distance between samples.
chem_data <- read_csv("../outputs/chem_data.csv")

#add an "s" in front of sample name to make it easier ot code later.
chem_data$Sample <- sub("^", "s", chem_data$Sample)
#Reduce chem data down to only the chemical values
chem_mat <- chem_data[,6:45]
#calculate euclidean distances between all samples and make matrix
chem_eucl <- as.matrix(dist(chem_mat, method="euclidean"))
#rename col and rownames with sample names.
rownames(chem_eucl)<-chem_data$Sample
colnames(chem_eucl)<-chem_data$Sample

#make vector of euclidean distances for all the comparisons
chem_dist <- as.vector(chem_eucl[upper.tri(chem_eucl)])

#OG Kush
og_kush_vec <- chem_eucl["s10", "s325"]
#104.1071

#Enemy of state
enemy_vec <- chem_eucl["s21", "s323"]
#47.0522

#Haze
haze_vec <- chem_eucl["s24", "s25"]
#9.786631

#Kush
kush_vec <- chem_eucl["s28", "s324"]
#12.87174

#Lemon Haze
chem_eucl["s5", "s7"]
lemon_vec <- chem_eucl["s5", "s7"]
#49.40912


#santa maria
chem_eucl["s11", "s328"]
santa_vec <- chem_eucl["s11", "s328"]
#20.07088

#amnesia matrix of distance between pairs
amnesia <- chem_eucl[c("s4", "s6", "s27", "s29", "s35", "s327", "s329", "s330"), c("s4", "s6", "s27", "s29", "s35", "s327", "s329", "s330")]
#make a vector of the amnesia matrix values
amnesia_vec <- as.vector(amnesia[upper.tri(amnesia)])
#calculate the median
median(amnesia_vec)
#median dist: 32.30729

#amnesia haze
amnesia_haze <- chem_eucl[c("s2", "s18", "s19", "s20", "s26", "s30", "s31", "s32", "s34"),c("s2", "s18", "s19", "s20", "s26", "s30", "s31", "s32", "s34")]
amnesia_haze_vec <- as.vector(amnesia_haze[upper.tri(amnesia_haze)])
median(amnesia_haze_vec)
#median 18.70257

#northern lights
northern <- chem_eucl[c("s23", "s33", "s320", "s1", "s3", "s8"), c("s23", "s33", "s320", "s1", "s3", "s8")]
northern_vec <- as.vector(northern[upper.tri(northern)])
median(northern_vec)
#median 43.47785

#turn into data frame
chem_dist <- as.data.frame(chem_dist) 

#make plot with median distance amongst the same clones

chem_clones <- ggplot(data=chem_dist, aes(chem_dist))+
  geom_histogram(bins=100, size = 0.8)+
  geom_vline(xintercept=49.40912, color="#F0E442", alpha=0.7)+ #lemon haze,yellow 
  geom_vline(xintercept=43.47785, color="#E69F00", alpha=0.7)+ #northern lights, light orange
  geom_vline(xintercept=20.07088, color="#56B4E9", alpha=0.7)+ #santa maria, light blue
  geom_vline(xintercept=18.70257, color="#009E73", alpha=0.7)+ #amnesia haze, dark green
  geom_vline(xintercept=32.30729, color="#999999", alpha=0.7)+ #amnesia,grey 
  geom_vline(xintercept=47.0522, color="#0072B2", alpha=0.7)+ #enemy of state, dark blue
  geom_vline(xintercept=12.87174, color="#D55E00", alpha=0.7)+ #kush, orange
  geom_vline(xintercept=104.1071, color="#CC79A7", alpha=0.7)+ #OG kush, pink
  geom_vline(xintercept=9.786631, color="#7570B3", alpha=0.7)+ #haze, purple
  labs(x="Euclidean distance", y="Count")

clones <- c(santa_vec, og_kush_vec, northern_vec, lemon_vec, kush_vec, haze_vec, enemy_vec, amnesia_haze_vec, amnesia_vec)


```

There are `r sum(chem_dist$chem_dist < 2.561392)`  pairwise comparisons between two samples with different labels that are more chemically similar to one another than any of the clones.

#Genetic distance between clones.

```{r gen_clones, message=FALSE, warning=FALSE}

#make a histogram showing the distribtions of genetic distances, highlight distances between clones with the same name.
gen_chem_data <- read_csv("../outputs/gen_chem_data.csv")
bedro_IBS <- read_table2("../data/20191209_bedrocan_gen_filtered.genome")
#The DST column contains the IBS value for the two samples being compared, we will calculate the inverse identity by state but doing 1-DST and we will label the new column inv_ibs.
bedro_IBS <- bedro_IBS %>% mutate(inv_ibs = 1-DST)
bedro_IBS <- bedro_IBS %>% select(FID1, FID2, inv_ibs)

#Pull out 
#OG kush: 10, 325
og_kush <- bedro_IBS[which(bedro_IBS$FID1 == 325 & bedro_IBS$FID2 == 10), "inv_ibs"]
og_kush <- as.vector(og_kush$inv_ibs)


#amnesia haze: 2, 20, 26, 31
amnesia_haze <- bedro_IBS[which(bedro_IBS$FID1 %in% c(2,20, 26, 31) & bedro_IBS$FID2 %in% c(2,20, 26, 31)), "inv_ibs"]
amnesia_haze <- as.vector(amnesia_haze$inv_ibs)
amnesia_haze_med <- median(amnesia_haze)

#amnesia: 6, 27, 329
amnesia <- bedro_IBS[which(bedro_IBS$FID1 %in% c(6, 27, 329) & bedro_IBS$FID2 %in% c(6, 27, 329)), "inv_ibs"]
amnesia <- as.vector(amnesia$inv_ibs)
amnesia_med <- median(amnesia)


#northern lights: 23, 33, 320, 1, 8
northern <- bedro_IBS[which(bedro_IBS$FID1 %in% c(23, 33, 320, 1, 8) & bedro_IBS$FID2 %in% c(23, 33, 320, 1, 8)), "inv_ibs"]
northern <- as.vector(northern$inv_ibs)
northern_med <- median(northern)

gen_clones <- ggplot(data=bedro_IBS, aes(bedro_IBS$`inv_ibs`))+
  geom_histogram(bins=100, size = 0.8)+
  geom_vline(xintercept=og_kush, color="#CC79A7", alpha=0.7)+ #OG kush, pink
  geom_vline(xintercept=amnesia_haze_med, color="#009E73", alpha=0.7)+ #amnesia haze, dark green
  geom_vline(xintercept=amnesia_med, color="#999999", alpha=0.7)+ #amnesia, grey
  geom_vline(xintercept=northern_med, color="#E69F00", alpha=0.7)+ #northern lights, light orange
  labs(x="Genetic distance (1-IBS)", y="Count")

clone_fig <- chem_clones / gen_clones + plot_annotation(tag_levels = 'a')

clone_fig
#ggsave("../figures/clone_fig.pdf", plot = clone_fig)
```

There are `r sum(bedro_IBS$inv_ibs < 0.078012)` pairs that are more genetic similar than of the clones.

#CBD vs. THC

```{r cbd_thc, message=FALSE, warning=FALSE}

#Plot CBD vs. THC
chem_data <- read_csv("../outputs/chem_data.csv")
ratio_cbd_thc <- dplyr::select(chem_data, "Sample", "thc_total", "CBD")

#get percent CBD and percent THC
ratio_cbd_thc <- mutate(ratio_cbd_thc, percent_thc = thc_total/10, percent_cbd = CBD/10)

cbd_thc <- ggplot(ratio_cbd_thc, aes(x=percent_cbd,y= percent_thc))+
  geom_point(size=1, alpha=0.7)+
  theme_bw()+
  labs(x="CBD (mg/g)", y="THC (mg/g)")+
  theme(panel.border = element_blank(), axis.text=element_text(colour = "black", size=10), axis.title=element_text(size = 10, face = "bold"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black", size = 0.2), plot.title = element_text(size = 9), legend.title = element_blank(), legend.position = "none")
#ggsave("../figures/cbd_thc.pdf", plot = cbd_thc)
cbd_thc

```

#Chemical correlation heatmap

```{r chem_cor, message=FALSE, warning=FALSE, include=FALSE}
#load chemical data
chem_data <- read_csv("../outputs/chem_data.csv")
chemicals <- chem_data[6:45]

chemicals[1:40] <- lapply(chemicals[1:40], as.character)
chemicals[1:40] <- lapply(chemicals[1:40], as.numeric)

#change from tbl to base dataframe.
chemicals <- as.data.frame(chemicals)

#create an empty matrix for correlations, make matrix with no data and just col and rows from final pheno table.
pairwise_pheno_correlations=matrix(,ncol(chemicals), ncol(chemicals))
rownames(pairwise_pheno_correlations)=colnames(chemicals)
colnames(pairwise_pheno_correlations)=colnames(chemicals)
#matrix for pvalues.
pairwise_pheno_correlations_pval=matrix(,ncol(chemicals), ncol(chemicals))
rownames(pairwise_pheno_correlations_pval)=colnames(chemicals)
colnames(pairwise_pheno_correlations_pval)=colnames(chemicals)

#Treat the data as all quantitative data and run pearson's correlation.
for (i in 1:ncol(chemicals)) {
  phenoname_x = colnames(chemicals)[i]
  for (j in 1:ncol(chemicals)) {
    phenoname_y = colnames(chemicals)[j]
    pairwise_pheno_correlations[j,i]=cor.test(chemicals[,i], chemicals[,j])$estimate
    pairwise_pheno_correlations_pval[j,i]= cor.test(chemicals[,i], chemicals[,j])$p.value
  }
}

#bonferonni correct the pval matrix
pairwise_pheno_correlations_pval[upper.tri(pairwise_pheno_correlations_pval)] = p.adjust(pairwise_pheno_correlations_pval[upper.tri(pairwise_pheno_correlations_pval)], method = "bonferroni")
pairwise_pheno_correlations_pval[lower.tri(pairwise_pheno_correlations_pval)] = p.adjust(pairwise_pheno_correlations_pval[lower.tri(pairwise_pheno_correlations_pval)], method = "bonferroni")

#for the supplement save the pval and rho matrices.
write.table(pairwise_pheno_correlations_pval, file = "../outputs/heatmap_pvals_r.csv", quote = F, sep = ",", row.names = F)
write.table(pairwise_pheno_correlations, file = "../outputs/heatmap_r.csv", quote = F, sep = ",", row.names = F)



#make the heat map/
#Set colour, make colour scale a gradient.
hmcol = colorRampPalette(brewer.pal(9, "BrBG"))

#Get one half of correlation square (aka one triangle). Set lower to NA.
pairwise_pheno_correlations[lower.tri(pairwise_pheno_correlations, diag = F)] <- NA

#make heat map for correlations without key.
pdf("../figures/chem_heatmap.pdf")
mar.default <- c(0,0,0,0)
par(mar=mar.default, cex = 0.5)
heatmap.2(pairwise_pheno_correlations,
          col = hmcol(1000),
          key=T,
          keysize = 0.5,
          symm = T,
          Colv = F,
          Rowv = F,
          scale = "none",
          trace="none",
          dendrogram = "none",
          density.info = "none",
          symbreaks = T,
          colsep = c(0:40),
          rowsep = c(0:40),
          sepcolor = "white",
          sepwidth = c(0.05, 0.05),
          symkey = T,
          margin = c(12,12)
)
dev.off()


#heat map for pvals of correlations.
#Get one half of correltaion square (aka one triangle). Set lower to NA.
pairwise_pheno_correlations_pval[upper.tri(pairwise_pheno_correlations_pval)] <- NA

#Pick colours.
hmcol_p = rev(brewer.pal(3, "YlGn"))
hmcol_p[3] = "#F7F7F7"

pdf("../figures/chem_heatmap_pval.pdf", width = 8, height = 8)
mar.default <- c(0,0,0,0)
par(mar=mar.default, cex = 0.5)
heatmap.2(pairwise_pheno_correlations_pval,
          col = hmcol_p,
          key=T,
          keysize = 0.5,
          symm = T,
          Colv = F,
          Rowv = F,
          trace="none",
          dendrogram = "none",
          density.info = "none",
          breaks = c(0, 0.001, 0.05, 1),
          revC = F,
          symbreaks = T,
          colsep = c(0:40),
          rowsep = c(0:40),
          sepcolor = "white",
          sepwidth = c(0.05, 0.05),
          symkey = T,
          margin = c(12, 12)
)
dev.off()


```

